### 2011-10-15 doc/manuscript.tex
Just an update.

% We approach the project of streptococcus recombination and gene transfer from
% the angle of a resource paper. 
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in}
\setlength{\parskip}{\bigskipamount}
\setlength{\parindent}{0pt}
\usepackage{url}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[authoryear]{natbib}
\usepackage{longtable}
\doublespacing

@@

\subsubsection*{Version: 0.67 (Oct 7, 2011) \clearpage{}}

@@

As sequencing costs plummet \citep{Mardis2011}, bacterial genome sequencing
produces individual genomes of high quality even within a bacterial species
\citep{Tettelin2009a}.  Studies on horizontal gene transfer in prokaryotes have
used genes in very large sets of species \citep{Koonin2001} from a great deal of
genomes available.  Although this type of study gave us insights into deep
evolutionary histories of prokaryotes, with multiple individual genomes per
species we now can go beyond the rough broad study on gene transfer towards
fine-scaled history of gene transfer between very closely related prokaryotes.
For instance, \citet{Liu2009} studied gene transfer events in
\textit{Lactobacillus bulgaricus} and \textit{Streptococcus thermophilus} using
the compositional approach of \citet{Karlin2001}. \citet{Luo2011} investigated
gene transfer events in \textit{Eschericha Coli} using the embedded quartet
decomposition due to \citet{Zhaxybayeva2006}.  \citet{Hamady2006} combined
phylogenetic and compositional approaches to detecting horizontal gene transfer.
\citet{Caro-Quintero2011} studied a number of \textit{Shewanella baltica}
genomes to identify the genetic elements that enabled the species to adapt to
redox gradients.  Although several complete genomes from SPY have been
available, \citet{Suzuki2011} just recently added one more complete genome of
SDE to the set of a previously unique SDE genome \citep{Shimomura2011}. These
multiple SDE genomes prompted us to study genome-wide gene flows between SPY and
SDE that afflicted human beings around the globe.

@@

In short, the complete genomes of the two spcies of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer. Virulence genes were found to be more associated
non-recombining horizontal gene transfer than non-virulence genes.  We compared
both of the results in recombining gene transfer events to show that the
parimony-based method inferred recombining gene transfer events in biological
functional categories similar to those inferred with the model-based approach.
This was the first attempt as far as we know of systematically analyzing SPY and
SDE in genome-wide scale to detect gene transfer events not only in the core
genome but also in the pan genome, which made the analysis complete in terms of
the amount of used genome data. Our analysis of core genomes also showed for the
first time that the net directionality of recombining gene transfer from SPY to
SDE was stronger than the reverse direction.

@@

There are 105 possible rooted labeled topologies relating five taxa.  We expected the
dominant topology across sampled genealogies to be concordant with that of the
species tree.  Indeed, two thirds of the local gene tree topologies (see section
\textbf{Materials and Methods}) along the alignment blocks were the same as the
species tree, which allowed us to be more confident of the species tree inferred
using the alignment blocks (Table S\ref{tab:Gene-tree-topologies}).  Other
frequently sampled topologies were expected to be potentially informative about
common recombination events.  The second most frequent topology (9.3\%) was one
where SDD was the outgroup, and SPY and SDE were sister clades. Therefore, SPY
and SDE appeared be more closely related with each other than each to SDD's
genome in about tenth of the core genome. The other topologies had less than 4\%
of proportions.

@@

The heat map of Figure \ref{fig:Heatmap-of-recombination} (see Tables
S\ref{tab:heatmap} and S\ref{tab:obsheatmap} for the heat map values and the
observed number of recombinant edges) shows the logarithm base 2 ratios of the
\textit{a posteriori} number of recombinant edges for a pair of donor and
recipient species tree branches and their corresponding expected values given
the species tree (Figure S\ref{fig:tree5}) and the three population parameters
(Table S\ref{tab:three}).  Because recombinant edges from any species tree
branch to the root branch are impossible (see Figure S\ref{fig:clonalorigin}),
the column for the root as recipient was colored green, indicating invalid
recombinant edges in Figure \ref{fig:Heatmap-of-recombination}.  
Similarly, all of the cells for which a recipient branch was
older than the donor, and the donor and recipient branches did not share any contemporary time were
colored green. The diagonal cells were colored gray, indicating no significant
number of recombinant edges, because the observed and expected numbers of
recombinant edges were similar. As recombinant edges that start from and end in
the same species tree branch would not change the species tree topology, which
might not
change the likelihood much, the cells would be colored gray.  Because SDD is a
strictly veterinary pathogenic bacteria, we did not expect that SDD received
much gene transfer from either SPY1 or SPY2: the logarithm base 2 ratio from SPY1 to
SDD was -2.20, and that from SPY2 to SDD was -1.97.  On the contrary, the
recombing gene transfer from SDE1 to SDD was 0.756, and that from SDE2 to SDD
was 0.698. This relatively higher recombining gene transfer from two SDE
individuals to SDD compared with that from two SPY individuals to SDD reflected
the fact
that the SDD was more closely related to SDE than to SPY in most of the genomes,
and the genomic organization between SDD and SDE1/SDE2 were much similar
(Figure S\ref{fig:mauve}).  However, the gene transfer from SDD to either SDE1
or SDE2 was less frequent than the case of reverse direction; SDD effectively
served as a recipient rather than a donor. It was not unclear why SDD received
more genes from SDE1 or SDE2 than did SDE1 or SDE2 from SDD.

Because it was expected that the recombining gene transfer within a clade would
be more frequent than that between clades, not surprisingly the recombining gene
transfer logarithm base 2 ratios between SDE1 and SDE2 (1.17 and 1.20) were about
twice as large as those from SDE1 and SDE2 to SPY1 and SPY2 (0.67, 0.66, 0.60,
and 0.39).  The intensities of recombining gene transfer from external branches
of SDE1 and SDE2 to external branches of SPY1 and SPY2 was similar to those from
SDE1 or SDE2 to SDD (0.76 and 0.70).  Therefore, either SPY1 or SPY2, and SDD as
recipient appeared to be no different from the view point of SDE1 or SDE2 as
donor since the similar values of recombining gene transfer intensity from SDE1
or SDE2 to one of the other three external branches. Surprisingly, the recombining
gene transfer logarithm base 2 ratios between SPY1 and SPY2 (1.06 and 1.07) was
comparable with, yet even less than those from SPY1 and SPY2 to SDE1 and SDE2
(1.25, 1.25, 1.17, and 1.29).  The recombining gene transfer from a clade SPY1 
or SPY2 to a different clade SDE1 or SDE2 seemed to be effectively no different
from that within the clade of SPY1 and SPY2.  The two external branches of SDE
received genes from either SPY1 or SPY2 via recombining gene transfer as if SDE1
and SDE2 were in the same clade as SPY1 and SPY2.  This was more surprising
because the bacterial recombination between SPY and SDE seemed to happen more
often than that between SDE and SDD even though the genome configurations of SDE
and SDD were more similar than each to SPY.  However, this result could be
concordant with the fact that SPY and SDE shared the same ecological niche
despite their dissimilar genomic organization.  The gene transfer logarithm base 2
ratio from SPY1 and SPY2 to SDE branch was also larger than any type of
recombining gene transfer: 1.88 and 1.94, respectively.  Additionally we found
58 genes for which a recombinant edge of certain types were sampled with
probability at least 0.9 by focusing on specific types of recombinant edges
(Table S\ref{tab:genes-transfer}).  Many of the genes in the list were
transferred from SPY to SDE rather than the reverse direction overall.  All
things considered, this epitomized that the recombining gene transfer was biased
toward the net direction from SPY to SDE.

@@

In addition to these biological categories, we also found several aminoacyl tRNA
synthetases (GO Term translation GO:0006412) in association with recombining
gene transfer.  In the parsimony-based approach the term translation was top
ranked (Table S\ref{tab:go-events-recombining}).  They had been a focus of gene
transfer in bacterial species \citep{Woese2000}.  Some of the functional
categories were unexpected; functional categories concerning ribosomes were
found to be associated with recombining gene transfer.  Because genes related
with ribosomes were considered to be housekeeping genes, gene transfer in which
would be rare, it was unexpected that ribosome related genes were found to be
associated with recombining gene transfer.  Because it was a bizarre result, we
investigated the association by considering different recombinant edge types. We
noted that two different classes of recombinant edge types: one that changes
species tree topologies, and another that does not. Because recombinant edges
that did not change species tree topology might be difficult to infer reliably,
considering those recombinant edges in computing recombinantion intensity might
lead to dubious results.  Ribosome related functional categories were found to
be top ranked if we took those recombinant edges that did not change species
tree topology (Tables S\ref{tab:functional-notopology} and
S\ref{tab:functional-all}).  In Table \ref{tab:functional} ribosome related
functional categories were not top ranked.  Therefore, high recombination
intensity in ribosome related genes were likely to be attributed to
topology-not-changing recombinant edges. We would be cautious of recombining
transfer of genes related with ribosome related functions.

@@

1. Topology-not-changing recombinant edge was unreliably estimated compared with
topology-changing recombinant edge.
2. Balanced case
3. Unbalanced case

We conducted experiments on simulation data to check the accuracy of estimates
obtained by the model-based approach in a scenario similar to one from the real 
data set.

While populaiton mutation and
recombination parameters were accurately estimated, the recombinant tract length
was over-estimated  (Table S\ref{tab:sim-three-population-parameters}).  
Recombinant tract lengths were reported to be difficult to estimate from relatively
short alignment blocks \citep{Didelot2010}.

Figure S\ref{fig:h1} shows the observed count of each recombinant
edge type against the expected count; the estimated numbers appeared to be
slightly over-estimated, but in correlation with the true numbers.

We also evaluated the accuracy of the statistics estimated in the second stage
of \texttt{ClonalOrigin}.

The number of recombinant edges for each pair of
recombinant tree branches was correlated with the true value in general while
there were slight over-estimation (Figure \ref{fig:h3}).

However, this
comparison between inferrence from simulated data and that of real data set
suggested that the directional difference between SPY and SDE was due to neither
lack of power of \texttt{ClonalOrigin} nor hidden bias thereof.

Recombination
intensities that were measured on genes also showed rough correlation (Pearson's
correlation coefficient of 0.63) of the estimated and true values (Figure
S\ref{fig:ri1}).

Estimated values appeared to be driven somewhat by the prior
implied by the population parameter setting, explaining why lower values were
over-estimated and higher values were under-estimated.

Nonetheless, the
correlation indicated that the measure of recombination intensity used in our
functional gene category served as an adequate indicator of actual intensity of
recombination events along genes.

@@

We present a short summary of the statistical method developed by
\citet{Didelot2007} and \citet{Didelot2010} for studying bacterial
recombination. A detailed account was found in the two papers and in the user
guide available at \url{http://code.google.com/p/clonalorigin}.  We used
\texttt{progressiveMauve} v2.3.1 \citep{Darling2004,Darling2010} with default
options to attempt to align the six genomes, which led to too small proportion
of core genomes. We had to remove the SEE sample from the genome alignment
although we kept it in the parsimony-based approach.  We then used
\texttt{stripSubsetLCBs}, a program of the \texttt{ClonalOrigin} package, and
identified 276 blocks of sequence alignments with length threshold of 1500 base
pairs.  We excluded two of these blocks in a subsequent analysis because they
contained long consecutive gaps.  We applied \texttt{ClonalFrame} v1.1 available
at \url{http://www.xavierdidelot.xtreemhost.com} to the remaining 274 blocks,
and estimated the species tree relating the five genomes. \texttt{ClonalFrame}
also estimates the ratio between recombination and mutation rates.
We used a single Markov chain Monte Carlo (MCMC) with 10000 burn-in generations
and 10000 sampling generations to obtain the tree estimate with branch lengths,
and additionally seven independent MCMC runs to validate adequate convergence of
the first run.

@@

\subsection{Simulation Study}
We performed three sets of simulations.  In the first simulation, we wished to
show that \texttt{ClonalOrigin} could infer population mutation rate, tract
length, and recombination rate with reasonable accuracy. In the second
simulation, we wished to check whether recombinant edges were recovered from
simulated data with recombinant trees generated from the prior. The third
simulation was conducted using simulated data with recombinant trees sampled the
posterior distribution.  We used a simulation feature of \texttt{ClonalOrigin}
to simulate the process of mutation and recombination across the five species
(see section \textbf{Supplementary Information} for detail).

@@

We first sampled ten recombinant trees given the mutation rate, recombination
rate, and tract length (Table S\ref{tab:three}) and the species tree (Figure
S\ref{fig:tree5}), and then created ten independent simulated data sets with
the Jukes-Cantor model for DNA sequence evolution using each of the recombinant
trees. This consisted of 100 replicates of data sets.  The recombinant trees
were created from the prior distribution and recombinant edges were added to the
species tree with no particular directions of recombining gene transfers between
species tree branches.  We computed the ratio of the mean number of recombinant
edges for a pair of donor and recipient species tree branches over the simulated
10 recombinant trees against the median of recombinant edges estimated from the
10 simulated data sets, and then took the logarithm of the ratio to base 2. We
also computed 5\% and 95\% quantiles to find the 90\% quantile range for the log
base 2 ratio.  We expect that the log base 2 ratio would near zero. We also
wished to see if how the recombinant edge count estimates from the real data set
deviate from the prior. Instead of using the number of recombinant edges from
the ten recombinant trees, we used the numbers of recombinant edges that were
estimated from the real data.

@@

We used 100 recombinant trees sampled from the posterior distribution
of the recombiant tree instead of using simulated recombinant trees.
We created one simulated data set for each of the 100 recombinant trees with
the Jukes-Cantor model for DNA sequence evolution 
using the mutation rate estimate (Table S\ref{tab:three}).
This consisted of 100 replicates of data sets.  
We follow the same computation of the ratio as in the simulation 2 with 
the numbers of recombinant edges estimated from the real data
as numerators.

@@

@@

\section*{Tables}

\begin{table}
\caption{\label{tab:functional}Functional categories asscoiated with 
recombination intensity based on topology changing recombinant edge types.}
\noindent \begin{centering}
\begin{tabular}{cccl}
\hline 
P-Value & Count & GO term & Description\tabularnewline
\hline 
1.731e-06 & 38 & GO:0009401 & phosphoenolpyruvate-dependent sugar phosphotransferase system\tabularnewline
2.775e-06 & 33 & GO:0008982 & protein-N(PI)-phosphohistidine-sugar phosphotransferase activity\tabularnewline
4.581e-05 & 22 & GO:0006633 & fatty acid biosynthetic process\tabularnewline
0.0001248 & 20 & GO:0008610 & lipid biosynthetic process\tabularnewline
0.0002675 & 45 & GO:0008643 & carbohydrate transport\tabularnewline
0.000513 & 81 & GO:0006412 & translation\tabularnewline
0.0005617 & 13 & GO:0006814 & sodium ion transport\tabularnewline
0.001019 & 40 & GO:0030529 & ribonucleoprotein complex\tabularnewline
0.001203 & 12 & GO:0031402 & sodium ion binding\tabularnewline
0.00206 & 39 & GO:0003735 & structural constituent of ribosome\tabularnewline
0.002115 & 237 & GO:0008152 & metabolic process\tabularnewline
0.00215 & 26 & GO:0019843 & rRNA binding\tabularnewline
0.00287 & 371 & GO:0016740 & transferase activity\tabularnewline
0.003552 & 66 & GO:0005840 & ribosome\tabularnewline
\hline 
\end{tabular}
\par\end{centering}
\end{table}
\clearpage{}

%
\begin{table}
\caption{\label{tab:go-events}Functional categories asscoiated with 
non-recombining horizontal gene transfer.}
\noindent \begin{centering}
\begin{tabular}{cccl}
\hline 
P-Value & Count & GO term & Description \\
\hline 
7e-61 &  67 & GO:0006313 & transposition, DNA-mediated\\
3e-58 &  59 & GO:0004803 & transposase activity\\
1e-30 &  52 & GO:0015074 & DNA integration\\
1e-28 &  23 & GO:0032196 & transposition\\
1e-13 & 527 & GO:0003677 & DNA binding\\
2e-12 &  86 & GO:0006310 & DNA recombination\\
1e-11 & 197 & GO:0003676 & nucleic acid binding\\
3e-09 &  36 & GO:0016987 & sigma factor activity\\
3e-09 &  36 & GO:0006352 & transcription initiation\\
3e-08 & 142 & GO:0043565 & sequence-specific DNA binding\\
\hline 
\end{tabular}
\par\end{centering}
\end{table}
\clearpage{}

\section*{Figures\clearpage{}}

@@

%=============================================================================
% Figure S12: 
\begin{figure}

@@

\caption{\label{fig:h3}Numer of recombination events. The diagonal dashed
lines are with slope of 1 and intercept of 0. For each order
pairs of branches of a reference species tree the true number of recombination
events is computed as the mean of the 10 replicates of recombinant
trees from the simulation. With ten estimates of each replicate total
number of 100 estimats are shown as intervals of 90\% of quantile
range, or the lower value of the interval is the 5\% quantile, and
the upper one is the 95\% quantile. }
\end{figure}
\clearpage{}%
%=============================================================================

@@

\section*{Supplementary Tables}
\clearpage{}

>>>>

\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in}
\setlength{\parskip}{\bigskipamount}
\setlength{\parindent}{0pt}
\usepackage{url}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[authoryear]{natbib}
\usepackage{longtable}
\doublespacing

@@

\subsubsection*{Version: 0.68 (Oct 15, 2011) \clearpage{}}

@@

As sequencing costs plummet \citep{Mardis2011}, bacterial genome sequencing
produces individual genomes of high quality even within a bacterial species
\citep{Tettelin2009a}.  Studies on horizontal gene transfer in prokaryotes have
used genes in very large sets of species \citep{Koonin2001} from a great deal of
genomes available.  Although this type of study gave us insights into deep
evolutionary histories of prokaryotes, with multiple individual genomes per
species we now can go beyond such rough broad study on gene transfer towards
fine-scaled history of gene transfer between very closely related prokaryotes.
For instance, \citet{Liu2009} studied gene transfer events in
\textit{Lactobacillus bulgaricus} and \textit{Streptococcus thermophilus} using
the compositional approach of \citet{Karlin2001}. \citet{Luo2011} investigated
gene transfer events in \textit{Eschericha Coli} using the embedded quartet
decomposition due to \citet{Zhaxybayeva2006}.  \citet{Hamady2006} combined
phylogenetic and compositional approaches to detecting horizontal gene transfer.
\citet{Caro-Quintero2011} studied a number of \textit{Shewanella baltica}
genomes to identify the genetic elements that enabled the species to adapt to
redox gradients.  Although several complete genomes from SPY have been
available, \citet{Suzuki2011} just recently added one more complete genome of
SDE to the set of a previously unique SDE genome \citep{Shimomura2011}. These
multiple SDE genomes prompted us to study genome-wide gene flows between SPY and
SDE that afflicted human beings around the globe.

@@

In short, the complete genomes of the two species of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer. Virulence genes were found to be more associated
non-recombining horizontal gene transfer than non-virulence genes.  We compared
both of the results in recombining gene transfer events to show that the
parsimony-based method inferred recombining gene transfer events in biological
functional categories similar to those inferred with the model-based approach.
This was the first attempt as far as we know of systematically analyzing SPY and
SDE in genome-wide scale to detect gene transfer events not only in the core
genome but also in the pan genome, which made the analysis complete in terms of
the amount of used genome data. Our analysis of core genomes also showed for the
first time that the net directionality of recombining gene transfer from SPY to
SDE was stronger than the reverse direction.

@@

There are 105 possible rooted labeled topologies relating five taxa.  We expected the
dominant topology across sampled genealogies to be concordant with that of the
species tree.  Indeed, two thirds of the local gene tree topologies (see section
\textbf{Materials and Methods}) along the alignment blocks were the same as the
species tree, which allowed us to be more confident of the species tree inferred
using the alignment blocks (Table S\ref{tab:Gene-tree-topologies}).  Other
frequently sampled topologies were expected to be potentially informative about
common recombination events.  The second most frequent topology (9.3\%) was one
where SDD was the outgroup, and SPY and SDE were sister clades. Therefore, SPY
and SDE appeared to be more closely related with each other than each to SDD's
genome in about tenth of the core genome. The other topologies had less than 4\%
of proportions.

@@

The heat map of Figure \ref{fig:Heatmap-of-recombination} (see Tables
S\ref{tab:heatmap} and S\ref{tab:obsheatmap} for the heat map values and the
observed number of recombinant edges) shows the logarithm base 2 fractions of the
\textit{a posteriori} number of recombinant edges for a pair of donor and
recipient species tree branches and their corresponding expected values given
the species tree (Figure S\ref{fig:tree5}) and the three population parameters
(Table S\ref{tab:three}).  Because recombinant edges from any species tree
branch to the root branch are impossible (see Figure S\ref{fig:clonalorigin}),
the column for the root as recipient was colored green, indicating invalid
recombinant edges in Figure \ref{fig:Heatmap-of-recombination}.  
Similarly, all of the cells for which a recipient branch was
older than the donor, and the donor and recipient branches did not share any contemporary time were
colored green. The diagonal cells were colored gray, indicating no significant
number of recombinant edges, because the observed and expected numbers of
recombinant edges were similar. As recombinant edges that start from and end in
the same species tree branch would not change the species tree topology, which
might not
change the likelihood much, the cells would be colored gray.  Because SDD is a
strictly veterinary pathogenic bacteria, we did not expect that SDD received
much gene transfer from either SPY1 or SPY2: the logarithm base 2 fraction from SPY1 to
SDD was -2.20, and that from SPY2 to SDD was -1.97.  On the contrary, the
recombing gene transfer from SDE1 to SDD was 0.756, and that from SDE2 to SDD
was 0.698. This relatively higher recombining gene transfer from two SDE
individuals to SDD compared with that from two SPY individuals to SDD reflected
the fact
that the SDD was more closely related to SDE than to SPY in most of the genomes,
and the genomic organization between SDD and SDE1/SDE2 were much similar
(Figure S\ref{fig:mauve}).  However, the gene transfer from SDD to either SDE1
or SDE2 was less frequent than the case of reverse direction; SDD effectively
served as a recipient rather than a donor. It was unclear why SDD received
more genes from SDE1 or SDE2 than did SDE1 or SDE2 from SDD.

Because it was expected that the recombining gene transfer within a clade would
be more frequent than that between clades, not surprisingly the recombining gene
transfer logarithm base 2 fractions between SDE1 and SDE2 (1.17 and 1.20) were about
twice as large as those from SDE1 and SDE2 to SPY1 and SPY2 (0.67, 0.66, 0.60,
and 0.39).  The intensities of recombining gene transfer from external branches
of SDE1 and SDE2 to external branches of SPY1 and SPY2 was similar to those from
SDE1 or SDE2 to SDD (0.76 and 0.70).  Therefore, either SPY1 or SPY2, and SDD as
recipient appeared to be no different from the view point of SDE1 or SDE2 as
donor since the similar values of recombining gene transfer intensity from SDE1
or SDE2 to one of the other three external branches. Surprisingly, the recombining
gene transfer logarithm base 2 fractions between SPY1 and SPY2 (1.06 and 1.07) was
comparable with, yet even less than those from SPY1 and SPY2 to SDE1 and SDE2
(1.25, 1.25, 1.17, and 1.29).  The recombining gene transfer from a clade SPY1 
or SPY2 to a different clade SDE1 or SDE2 seemed to be effectively no different
from that within the clade of SPY1 and SPY2.  The two external branches of SDE
received genes from either SPY1 or SPY2 via recombining gene transfer as if SDE1
and SDE2 were in the same clade as SPY1 and SPY2.  This was more surprising
because the bacterial recombination between SPY and SDE seemed to happen more
often than that between SDE and SDD even though the genome configurations of SDE
and SDD were more similar than each to SPY.  However, this result could be
concordant with the fact that SPY and SDE shared the same ecological niche
despite their dissimilar genomic organization.  The gene transfer logarithm base 2
fraction from SPY1 and SPY2 to SDE branch was also larger than any type of
recombining gene transfer: 1.88 and 1.94, respectively.  Additionally we found
58 genes for which a recombinant edge of certain types were sampled with
probability at least 0.9 by focusing on specific types of recombinant edges
(Table S\ref{tab:genes-transfer}).  Many of the genes in the list were
transferred from SPY to SDE rather than the reverse direction overall.  All
things considered, this epitomized that the recombining gene transfer was biased
toward the net direction from SPY to SDE.

@@

In addition to these biological categories, we also found several aminoacyl tRNA
synthetases (GO Term translation GO:0006412) in association with recombining
gene transfer.  In the parsimony-based approach the term translation was top
ranked (Table S\ref{tab:go-events-recombining}).  They had been a focus of gene
transfer in bacterial species \citep{Woese2000}.  Some of the functional
categories were unexpected; functional categories concerning ribosomes were
found to be associated with recombining gene transfer.  Because genes related
with ribosomes were considered to be housekeeping genes, gene transfer in which
would be rare, it was unexpected that ribosome related genes were found to be
associated with recombining gene transfer.  Because it was a bizarre result, we
investigated the association by considering different recombinant edge types. We
noted that two different classes of recombinant edge types: one that changes
species tree topologies, and another that does not. Because recombinant edges
that did not change species tree topology might be difficult to infer reliably,
considering those recombinant edges in computing recombination intensity might
lead to dubious results.  Ribosome related functional categories were found to
be top ranked if we took those recombinant edges that did not change species
tree topology (Tables S\ref{tab:functional-notopology} and
S\ref{tab:functional-all}).  In Table \ref{tab:functional} ribosome related
functional categories were not top ranked.  Therefore, high recombination
intensity in ribosome related genes were likely to be attributed to
topology-not-changing recombinant edges. We would be cautious of recombining
transfer of genes related with ribosome related functions.

@@

We conducted experiments on simulation data to check the accuracy of estimates
obtained by the model-based approach in a scenario similar to the one from the
real data set.  Population mutation and recombination parameters were accurately
estimated (Table S\ref{tab:sim-three}).  Yet, the recombinant tract length was
over-estimated; recombinant tract lengths were reported to be difficult to
estimate from relatively short alignment blocks \citep{Didelot2010}.  Numbers of
recombinant edges from the simulated recombinant trees of Simulation 2 (see
\textbf{Supplementary Information} for the description of Simulation 2) were
compared with the corresponding numbers of recombinant edges that were estimated
from the simulated data set from Simulation 2 (see the solid-line intervals with
open circles in Figure \ref{fig:h3}).  We noted that tree-topology-changing
recombinant edges were more reliably recovered compared with
tree-topology-not-changing ones (Figure \ref{fig:h3}). All of the
tree-topology-not-changing recombinant edge types were over-estimated (Figure
\ref{fig:h3}B). The heat map of Figure \ref{fig:Heatmap-of-recombination} did
not show how confident of the values we could be. Comparison of numbers of
recombinant edges estimated from the real data set and the corresponding numbers
estimated from Simulation 2 showed the interval estimates for the heat map
values.  Dotted-line intervals from the red and blue colored cells in the heat
map were above and below the reference of the horizontal line at the zero on the
Y-axis, respectively (Figure \ref{fig:h3}).

for each recombinant edge type to assess the effect of posterior recombinant
trees samples with respect to that of prior recombinant tree samples. The ten
\textit{a priori} recombinant trees from Simulation 2 would have balanced
recombining gene transfer events between SPY and SDE with respect to the prior
expected number of recombinant edges. The posterior recombinant trees used in
Simulation 3 would have unbalanced recombining gene transfer between the two
species branches as we observed in the real data analysis. For significantly
strong gene flows such as ones from SPY to SDE a solid-line interval with a
closed circle would be different from its count part of the dotted-line
interval.  Figure \ref{fig:h3} showed that the recombining gene flow from SPY to
SDE was stronger enough to be nearer to the horizontal line at the zero on the
Y-axis. However, the recombining gene transfer from SDE to SPY appeared to be
too weak.  This comparison supported the biased directionality of recombining
gene transfer from SPY to SDE.

Recombination intensities that were measured on genes also showed rough
correlation (Pearson's correlation coefficient of 0.63) of the estimated and
true values (Figure S\ref{fig:ri1}).  Estimated values appeared to be driven
somewhat by the prior implied by the population parameter setting, explaining
why lower values were over-estimated and higher values were under-estimated.
Nonetheless, the correlation indicated that the measure of recombination
intensity used in our functional gene category served as an adequate indicator
of actual intensity of recombination events along genes.

@@

We present a short summary of the statistical method developed by
\citet{Didelot2007} and \citet{Didelot2010} for studying bacterial
recombination. A detailed account was found in the two papers and in the user
guide available at \url{http://code.google.com/p/clonalorigin}.  We used
\texttt{progressiveMauve} v2.3.1 \citep{Darling2004,Darling2010} with default
options to attempt to align the six genomes, which led to too small proportion
of core genomes. We had to remove the SEE sample from the genome alignment
although we kept it in the parsimony-based approach.  We then used
\texttt{stripSubsetLCBs}, a program of the \texttt{ClonalOrigin} package, and
identified 276 blocks of sequence alignments with length threshold of 1500 base
pairs.  We excluded two of these blocks in a subsequent analysis because they
contained long consecutive gaps.  We applied \texttt{ClonalFrame} v1.1 available
at \url{http://www.xavierdidelot.xtreemhost.com} to the remaining 274 blocks,
and estimated the species tree relating the five genomes. \texttt{ClonalFrame}
also estimates the fraction between recombination and mutation rates.
We used a single Markov chain Monte Carlo (MCMC) with 10000 burn-in generations
and 10000 sampling generations to obtain the tree estimate with branch lengths,
and additionally seven independent MCMC runs to validate adequate convergence of
the first run.

@@

\subsection{Simulation Study}

We performed three sets of simulations.  In the first simulation, we wished to
show that \texttt{ClonalOrigin} could infer population mutation rate, tract
length, and recombination rate with reasonable accuracy. In the second
simulation, we wished to check whether recombinant edges were recovered from
simulated data with recombinant trees generated from the prior. The third
simulation was conducted using simulated data with recombinant trees sampled the
posterior distribution.  We used a simulation feature of \texttt{ClonalOrigin}
to simulate the process of mutation and recombination across the five species
(see section \textbf{Supplementary Information} for detail).

@@

We first sampled ten recombinant trees given the mutation rate, recombination
rate, and tract length (Table S\ref{tab:three}) and the species tree (Figure
S\ref{fig:tree5}), and then created ten independent simulated data sets with
the Jukes-Cantor model for DNA sequence evolution using each of the recombinant
trees. This consisted of 100 replicates of data sets.  The recombinant trees
were created from the prior distribution and recombinant edges were added to the
species tree with no particular directions of recombining gene transfers between
species tree branches.  We computed the fraction of the mean number of recombinant
edges for a pair of donor and recipient species tree branches over the simulated
10 recombinant trees against the median of recombinant edges estimated from the
10 simulated data sets, and then took the logarithm of the fraction to base 2. We
also computed 5\% and 95\% quantiles to find the 90\% quantile range for the log
base 2 fraction.  We expect that the log base 2 fraction would near zero. We also
wished to see if how the recombinant edge count estimates from the real data set
deviate from the prior. Instead of using the number of recombinant edges from
the ten recombinant trees, we used the numbers of recombinant edges that were
estimated from the real data.

@@

We used 100 recombinant trees sampled from the posterior distribution
of the recombiant tree instead of using simulated recombinant trees.
We created one simulated data set for each of the 100 recombinant trees with
the Jukes-Cantor model for DNA sequence evolution 
using the mutation rate estimate (Table S\ref{tab:three}).
This consisted of 100 replicates of data sets.  
We follow the same computation of the fraction as in the simulation 2 with 
the numbers of recombinant edges estimated from the real data
as numerators.

@@

% \clearpage{}

% \section*{Tables}

@@

%=============================================================================
% Table 1. 
\begin{table}
\caption{\label{tab:functional}Functional categories asscoiated with 
recombination intensity based on topology changing recombinant edge types.}
\noindent \begin{centering}
\begin{tabular}{cccl}
\hline 
P-Value & Count & GO term & Description\tabularnewline
\hline 
1.731e-06 & 38 & GO:0009401 & phosphoenolpyruvate-dependent sugar phosphotransferase system\tabularnewline
2.775e-06 & 33 & GO:0008982 & protein-N(PI)-phosphohistidine-sugar phosphotransferase activity\tabularnewline
4.581e-05 & 22 & GO:0006633 & fatty acid biosynthetic process\tabularnewline
0.0001248 & 20 & GO:0008610 & lipid biosynthetic process\tabularnewline
0.0002675 & 45 & GO:0008643 & carbohydrate transport\tabularnewline
0.000513 & 81 & GO:0006412 & translation\tabularnewline
0.0005617 & 13 & GO:0006814 & sodium ion transport\tabularnewline
0.001019 & 40 & GO:0030529 & ribonucleoprotein complex\tabularnewline
0.001203 & 12 & GO:0031402 & sodium ion binding\tabularnewline
0.00206 & 39 & GO:0003735 & structural constituent of ribosome\tabularnewline
0.002115 & 237 & GO:0008152 & metabolic process\tabularnewline
0.00215 & 26 & GO:0019843 & rRNA binding\tabularnewline
0.00287 & 371 & GO:0016740 & transferase activity\tabularnewline
0.003552 & 66 & GO:0005840 & ribosome\tabularnewline
\hline 
\end{tabular}
\par\end{centering}
\end{table}
%\clearpage{}
%=============================================================================

%=============================================================================
% Table 2. 
\begin{table}
\caption{\label{tab:go-events}Functional categories asscoiated with 
non-recombining horizontal gene transfer.}
\noindent \begin{centering}
\begin{tabular}{cccl}
\hline 
P-Value & Count & GO term & Description \\
\hline 
7e-61 &  67 & GO:0006313 & transposition, DNA-mediated\\
3e-58 &  59 & GO:0004803 & transposase activity\\
1e-30 &  52 & GO:0015074 & DNA integration\\
1e-28 &  23 & GO:0032196 & transposition\\
1e-13 & 527 & GO:0003677 & DNA binding\\
2e-12 &  86 & GO:0006310 & DNA recombination\\
1e-11 & 197 & GO:0003676 & nucleic acid binding\\
3e-09 &  36 & GO:0016987 & sigma factor activity\\
3e-09 &  36 & GO:0006352 & transcription initiation\\
3e-08 & 142 & GO:0043565 & sequence-specific DNA binding\\
\hline 
\end{tabular}
\par\end{centering}
\end{table}
\clearpage{}
%=============================================================================

%\section*{Figures\clearpage{}}

@@

%=============================================================================
% Figure 4. 
\begin{figure}

@@

\caption{\label{fig:h3}
Simulations of the model-based approach for assessing the power of recovering
recombinant edges. Pairs of species tree branches are shown at x-axis labels.
Three intervals are displayed for each pair. Intervals with solid line with open
circle in the middle are for the logarithm of the fraction to base 2 of the average
number of recombinant edges from the ten simulated recombinant trees 
to the numbers of recombinant edges estimated from Simulation 2. 
Interverals with dotted line are for the logarithm of the fraction to base 2 of
the number of recombinant edges estimated from the real data
to the numbers of recombinant edges estimated from Simulation 2.
Intervals with solid line with closed circle in the middle are for the logarithm
of the fraction to base 2 of the number of recombinant edges estimated from the real data
to the numbers of recombinant edges estimated from Simulation 3. 
The horizontal lines are at the zero.
The intervals range from 5\% quantile to 95\% of quantile.}
\end{figure}
\clearpage{}%
%=============================================================================

@@

% \section*{Supplementary Tables}
% \clearpage{}
